---
title: "Using moment equations to understand heterogeneity"
output: pdf_document
---

## Definition

We want to write $Y_o(v;D)$ for T**y**pe, **o**rder, **v**ariable, and **D**omain.

We define total as follows: $T_o(v;D) = \int v(a)^o D(a) da$. Since we're mainly interested with susceptibility for now, $T_o$ is represents $T_o(.;S)$.

We define $M_i = T_i/T_0$. Then, $M_1$ is the mean susceptibility and $\kappa = \frac{M_2M_0}{M_1^2}-1$ is the squared coefficient of variance (CV).

## SI model

We have $\dot{S}(a) = - \Lambda \sigma(a) S(a)$. Integrating gives us $\dot{T}_0 = - \Lambda T_1$. More generally, we have $\dot{T}_i = - \Lambda T_{i+1}$. Given that $M_1$ is the mean susceptibility, we can also write:

$$
\dot{S} = - \Lambda M_1 S
$$

Using $M$ defined above, we also have the following equations: $\dot{M}_i = - \Lambda (M_{i+1} - M_i M_1)$.

Given that $M_2 = (1 + \kappa) M^2$ and assuming that $\kappa$ stays constant, we can integrate the equation above to obtain the following equation:

$M = \hat{M}S^\kappa$, where $\hat{M}$ is the mean susceptibility of the susceptible population at a disease free equilibrium.

### Definition - $\kappa_i = \frac{M_i}{M_{i-1}M_1}-1$

$$
\begin{aligned}
\dot{\kappa}_i &=\frac{M_{i-1}M_1 \dot{M}_i - M_i M_1 \dot{M}_{i-1} - M_i M_{i-1} \dot{M}_1}{M_{i-1}^2 M_1^2}\\
&=- \Lambda \frac{M_{i-1}M_1 (M_{i+1}-M_iM_1) - M_i M_1 (M_i-M_{i-1}M_1) - M_i M_{i-1} (M_2 - M_1^2) }{M_{i-1}^2 M_1^2}\\
&=- \Lambda \frac{M_{i-1}M_1 M_{i+1} - M_i M_1 M_i - M_i M_{i-1} M_2 + M_i M_{i-1} M_1^2 }{M_{i-1}^2 M_1^2}\\
&=- \Lambda \frac{(\kappa_{i+1}+1) M_i M_{i-1} M_1^2  - M_i M_1 M_i - (\kappa_2 +1) M_i M_{i-1} M_1^2 + M_i M_{i-1} M_1^2 }{M_{i-1}^2 M_1^2}\\
&=- \Lambda \frac{(\kappa_{i+1}+1) M_i M_{i-1} M_1 - (\kappa_i+1) M_{i-1} M_i M_1 - \kappa_2 M_i M_{i-1} M_1}{M_{i-1}^2 M_1}\\
&=- \Lambda \frac{(\kappa_{i+1}+1) M_i - (\kappa_i+1) M_i - \kappa_2 M_i}{M_{i-1}}\\
&=- \Lambda M_i \frac{\kappa_{i+1} - (\kappa_2 +\kappa_i)}{M_{i-1}}\\
&=- \Lambda M_1 (\kappa_i + 1) \{\kappa_{i+1} - (\kappa_2 +\kappa_i)\}\\
\end{aligned}
$$

How accurate is this?

```{r load, echo = FALSE, message= FALSE}
load("SI_sim.rda")
```

### Ex 1 - lognormal distribution

Linear equation:
$$r \approx (\hat{r}-2)S + 2$$

- Heterogeneous model 
- Approximated model: linear equation for $\kappa_3/\kappa_2$ coupled with approximated equation for $\kappa$
- Approximated model: constant $\kappa$

```{r plotComparison, fig.width = 15, fig.height=5}
source("plotFunctions.R")
name <- c("Heterogeneous", "Approximation 1", "Approximation 2")
plotSim(SI.list.lnorm, name)
```

## SIS model

Here is a simple SIS model: $\dot{S}(a) = \rho(N(a)-S(a)) - \Lambda \sigma(a) S(a)$ and $\dot{I} = \int \Lambda 
\sigma S(a) da - I$, where $N(a)$ is the initial distribution of the susceptible individuals in a disease free equilibrium. We are going to define $\bar{Y}_o = Y_o(v;N)$. For this model, we have $\dot{T}_i = \rho (\bar{T}_i - T_i) - \Lambda T_{i+1}$. Using chain rule, we can also get an equation for $M = M_1$:

$$
\begin{aligned}
\dot{M}_i &= \frac{\dot{T}_i}{T_0} - \frac{\dot{T_0} T_i}{T_0^2}\\
&= (\rho (\bar{T}_i/T_0 -M_i) - \Lambda M_{i+1}) - \rho(M_i \bar{T}_0/T_0 - M_i) + \Lambda M_i M_1\\
&= \rho (\bar{M}_i- M_i) \frac{\bar{T}_0}{T_0} - \Lambda (M_{i+1} - M_i M_1)\\
&= \rho (\bar{M}_i- M_i)/T_0 - \Lambda \kappa_{i+1} M_i M_1\\
\end{aligned}
$$

When $i = 1$, $\dot{M} = \rho (\bar{M}_1 - M_1)/T_0 - \Lambda \kappa M_1^2$.

### Derivative of $\kappa$

New variable: $\varphi_i = \kappa_i + 1$.

$$
\begin{aligned}
\dot{\varphi}&=\varphi(\frac{\dot{M}_2}{M_2} - 2 \frac{\dot{M}_1}{M_1})\\
&=\varphi(\frac{\rho(\bar{M}_2-M_2)/T_0-\Lambda \kappa_3 M_2 M_1}{M_2} - 2 \frac{\rho(\bar{M}_1-M_1)/T_0 - \Lambda \kappa_2 M_1^2}{M_1})\\
&=\varphi\{\frac{\rho}{T_0}(\frac{(\bar{M}_2-M_2)}{M_2} - 2 \frac{(\bar{M}_1-M_1)}{M_1}) -\Lambda M_1 (\varphi_3 - 2 \varphi + 1)\}\\
&=\varphi\{\frac{\rho}{T_0}(\frac{M_1\bar{M}_2+M_1M_2 - 2 M_2 \bar{M}_1}{M_2M_1}) -\Lambda M_1 (\varphi_3 - 2 \varphi + 1)\}\\
&=\varphi\{\frac{\rho}{T_0}(\frac{\bar{M}_2}{M_2}-2 \frac{\bar{M}_1}{M_1} + 1) -\Lambda M_1 (\varphi_3 - 2 \varphi + 1)\}\\
&=\varphi\{\frac{\rho}{T_0}(\frac{\bar{\varphi}}{\varphi}\frac{\bar{M}_1^2}{M_1^2}-2 \frac{\bar{M}_1}{M_1} + 1) -\Lambda M_1 (\varphi_3 - 2 \varphi + 1)\}\\
&=\varphi\{\frac{\rho}{T_0}(\frac{\bar{\varphi}\bar{M}_1^2 - 2\varphi\bar{M}_1M_1 + \varphi M_1^2}{\varphi M_1^2}) -\Lambda M_1 (\varphi_3 - 2 \varphi + 1)\}\\
&=\varphi\{\frac{\rho}{T_0}(\frac{(\bar{\varphi}-\varphi)\bar{M}_1^2 + \varphi (\bar{M}_1 - M_1)^2}{\varphi M_1^2}) -\Lambda M_1 (\varphi_3 - 2 \varphi + 1)\}\\
\end{aligned}
$$



```{r SIS, fig.width = 15, fig.height=5}
load("SIS_sim.rda")
plotSim(SIS.list.lnorm, name)

```

Approximation 1: 

- Constant $\kappa$
- $M = \hat{M}S^\kappa$

Approximation 2:

- $\dot{\varphi}=\varphi\{\frac{\rho}{T_0}(\frac{(\bar{\varphi}-\varphi)\bar{M}_1^2 + \varphi (\bar{M}_1 - M_1)^2}{\varphi M_1^2})$
- $\dot{M} = \rho (\bar{M}_1 - M_1)/T_0 - \Lambda \kappa M_1^2$

We should probably try another simulation with constant $\kappa$ coupled with the differential equation for $M$...